%!TEX TS-program = xelatex 

%!TEX encoding = UTF-8 Unicode
%
%  proposal
%
%  Created by Mark Eli Kalderon on 2015-09-30.
%  Copyright (c) 2015. All rights reserved.
%

\documentclass[12pt]{article} 

% Definitions
\newcommand\mykeywords{perception, sympathy, presence}
\newcommand\myauthor{Mark Eli Kalderon}

% Packages
\usepackage{geometry} \geometry{a4paper}

% XeTeX
\usepackage[cm-default]{fontspec}
\usepackage{xltxtra,xunicode}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\setmainfont{Hoefler Text}

% Bibliography
\usepackage[round]{natbib}

% PDF Stuff
\usepackage[plainpages=false, pdfpagelabels, bookmarksnumbered, backref, pdftitle={Parousia, Sympathy and Sensory Presentation}, pagebackref, pdfauthor={\myauthor}, pdfkeywords={\mykeywords}, xetex]{hyperref}

%%% BEGIN DOCUMENT
\begin{document}

% % Title Page
\author{\myauthor}
\title{Book Proposal\\
\emph{Parousia}\\
Sympathy and Sensory Presentation}
\date{}

\maketitle

% Layout Settings
\setlength{\parindent}{1em}

% Main Content

\section{Overview} % (fold)
\label{sec:overview}

The present essay is an unabashed exercise in historically informed, speculative metaphysics. Its aim is to gain insight into the nature of sensory presentation.

One of the fundamental issues dividing contemporary philosophers of perception is whether perception is presentational or representational in character. To claim that perception is presentational in character is to claim that it has a presentational element irreducible to whatever intentional or representational content it may have. So conceived, the object of perception is present in the awareness afforded by the perceptual experience and is thus a constituent of that experience. Representationalists deny that perception has such an irreducible presentational element, claiming, instead, that the object of perception is exhaustively specified by its intentional or representational content. If there is indeed a presentational element to perception, then, according to the representationalist, this is because sensory presentation is either reducible to the exercise of an intentional or representational capacity or otherwise essentially involves the exercise of such a capacity. There are two aspects of this debate. On the one hand, there are arguments on one side or the other urging that perception must be conceived in presentational or representational terms. One the other hand, there is a more positive, constructive aspect, where, taking for granted one’s preferred conception, one goes on to develop detailed theoretical accounts of perceptual experience.

Representationalists have been more active in this latter task. And unsurprisingly so. For suppose one took sensory presentation to be an indispensable aspect of perceptual experience and further held, in a Butlerian spirit, that it was reducible to no other thing. What positive account could one give of sensory presentation, so conceived? Since it is irreducible, no positive account could take the form of a reduction. So no causal or counterfactual conditions on sensory representations, understood independently of perception, could be jointly necessary and sufficient for the presentation, in sensory experience, of its object. One might specify the
relational features of presentation in sensory experience, but not much insight into the nature of sensory presentation is thereby gained. The tools of contemporary analytic metaphysics would seem not to leave one much to work with, at least in the present instance. So it can seem that if one maintains that perceptual experience involves an irreducible presentational element, all that one can do is press the negative point that sensory presentation, an indispensable element of perceptual experience, is reducible to no other thing.

I believe that perception has an irreducible presentational element. And yet I hoped to learn something positive about the metaphysics of sensory presentation. If there was, in fact, anything further to be learned, I could not limit myself to the tools of contemporary analytic metaphysics. The present metaphysics is historically informed, at least in part, as a result of looking for tools more adequate to the task at hand. 

Tactile metaphors for perception are primordial and persistent. What makes such tactile metaphors for perception, such as grasping and apprehension, apt? 

Grasping is at the center of a semantic field of tactile metaphors for sensory awareness loosely organized as modes of assimilation. I attempt to understand what, if anything, makes grasping an apt metaphor for sensory awareness more generally by undertaking a phenomenological investigation into grasping or enclosure understood as a mode of haptic perception. The idea is that if we better appreciate how grasping presents itself from within haptic experience, we will be in a better position to understand what, if anything, makes grasping an apt metaphor for perception generally. Moreover, in undertaking this phenomenological investigation we shall freely draw upon empirical and historical sources. Empirical psychology has a lot to teach us about the phenomenology of haptic experience. But so does the testimony of our respected predecessors and the puzzles that arise both within and without the \emph{endoxa}.

Moreover, there is reason why a phenomenological investigation into haptic experience whose ultimate aim is to uncover the aptness of tactile metaphors for perception generally should take the form of a conceptual genealogy. In looking at earlier occurrences of such metaphors, when they were more strongly etched in light and shadow, one can get a better sense of what made them live for these earlier thinkers and, by extension, a better sense of the power they continue to exercise over us.

The central claim of the present essay is that sensory presentation might be understood in terms of the operation of sympathy, construed not as fellow-feeling so much as the principle at work in fellow-feeling. Hence, sensory presentation is a kind of sympathetic presentation. This claim is first presented as a resolution of a puzzle about how grasping could be the presentation of something extra-somatic. The essay develops accounts of auditory and visual presentation in terms of the operation of sympathy and develops an account of perceptual objectivity on this basis as well.


% section overview (end)

\section{The Audience} % (fold)
\label{sec:the_audience}

The present essay is primarily addressed to philosophers of perception and anyone interested in the metaphysics of experience (as well as graduate students working in these fields). It may also be of interest to historians of philosophy working on these issues as it draws upon and develops some historical issues as they are relevant to the argument of the essay.

% section the_audience (end)

\section{The Competition} % (fold)
\label{sec:the_competition}

Only two books come to mind.

First David Smith's \emph{The Problem of Perception}. Smith argues that sensory presentation must be understood in terms of Fichte's notion of \emph{Anschloss}. The present essay, in effect, argues that this is a half-truth, and that the neo-Fichtean notion is better understood as explicable in terms of a precondition for the operation of sympathy as it occurs in sensory presentation.

The other comparable book would be Alva No\"{e}'s \emph{Varieties of Presence}. Like No\"{e}, I am taking seriously the metaphysics of sensory presentation while admitting that it reduces to no other thing. There are substantive differences, however. No\"{e}'s work is not historically informed in the way the present work is. Moreover, No\"{e}, at least by my lights, is a skeptic about sensory presentation in a way that I am not (that, I take it, is part of the force of the qualifier ``virtual'' and No\"{e}'s emphasis on perceptual availability in accounting for virtual presence).

% section the_competition (end)

\section{Professional Experience} % (fold)
\label{sec:professional_experience}

I am a professor of philosophy at University College London. I have written extensively on color and the philosophy of perception. Not only have I done so over a series of peer-reviewed journal articles and chapters \citep{Hilbert:2000on,Kalderon:2006tg,Kalderon:2008fk,Kalderon:2007mr,Kalderon:2010fj,Kalderon:2010fk,Kalderon:2011fk,Kalderon:2012fk}, but I have also written a monograph on Empedocles and Aristotle on color perception \citep{Kalderon:2015fr}.

% section professional_experience (end)

\section{Chapters} % (fold)
\label{sec:chapters}

The present essay consists of 6 chapters:

\begin{enumerate}
	\item \emph{Grasping} Discusses grasping as a mode of haptic presentation and raises a puzzle concerning the haptic presentation of the extra-somatic given the role of bodily awareness in haptic perception.
	\item \emph{Sympathy} Discusses responses to the puzzle and resolves it by accounting for haptic presentation as a species of sympathy.
	\item \emph{Sound} Defends the Wave Theory of Sound.
	\item \emph{Sources of Sound} Argues that the sources of sound are heard in or through the sounds they make and explains this in terms of sympathetic presentation.
	\item \emph{Vision} Extends the account of sensory presentation in terms of sympathy to visual presentation.
	\item \emph{Realism} Argues that sympathetic presentation may present how things are in themselves and that the phenomenal/noumenal distinction, on a certain understanding, collapses.
\end{enumerate}

% section chapters (end)

\section{Estimated Time of Delivery} % (fold)
\label{sec:estimated_time_of_delivery}

I presently have an advanced draft of the manuscript of approximately 80,000 words.

% section estimated_time_of_delivery (end)

% Bibligography
\bibliographystyle{plainnat}
\bibliography{Philosophy}

\end{document}
