%!TEX root = /Users/markelikalderon/Documents/Git/sympathy/perception.tex
\chapter{Sources of Sound} % (fold)
\label{cha:sources_of_sound}

\section{The Heideggerian Alternative} % (fold)
\label{sec:the_heideggerian_alternative}

On one understanding, the source of a sound may be a body that possesses the power of sounding, that is, the power to engage in a sound-generating activity. On another understanding, the source of a sound is simply the sound-generating activity, the event or process that generates the sound. Indeed, there may be sound-generating events or processes not involving the activity of bodies (though for the most part we shall ignore this possibility). Audible sources, the sources disclosed in auditory experience, are sound-generating events or processes. It is the audible activities of bodies, or at least sound-generating events or processes, that we hear and not the bodies themselves (though perhaps we may attend to these on the supposition that their audible activities constitute a dynamic aural image of them, chapter~\ref{sec:sounds_and_their_sources}). We hear sounds and their sources. What role do sounds play in affording the perceiver with auditory awareness of their sources?

The neo-Berkelean has an answer, ready to hand, that many find nearly irresistible. We hear the sources of sound by hearing the sounds they generate. We hear a body's audible activity by hearing the sound that activity generates. Hearing sounds afford the perceiver with auditory awareness of their sources since the immediate presentation of the sound in auditory experience constitutes, in a manner yet to be explained, the mediate presentation of its source.

I do not accept the neo-Berkelean answer. I believe that its central claims are at odds with the phenomenology of auditory experience. Instead, I shall refine and elaborate a Heideggerian account of the role that sounds play in affording the perceiver with auditory awareness of their distal sources (see \citealt{Leddington:2014aa} for a different defence of the Heideggerian alternative and the replies by \citealt{OCallaghan:2014aa} and \citealt{Nudds:2014aa}). We do not hear sources by hearing their sounds as the neo-Berkelean would have it. Rather we hear the sources of sound directly. In cases where a perceiver can hear the source of a sound, the call of a feral parrot, say, they are explicitly aware of the call and only implicitly aware of its sound. The application of Fulkerson's \citeyearpar{Fulkerson:2014ek} distinction between explicit and implicit awareness to Heidegger's \citeyearpar{Heidegger:1935uq} observation about audition is the first of the refinements. There is a sense in which we hear a source through, or in \citep{Leddington:2014aa}, the sound it produces. The sound they hear is a perceptual medium through which the audible activities of distal bodies are disclosed. And sympathy is the principle that makes possible the presentation of sources in auditory experience through the perceptual medium of sound. On the refined and elaborated Heideggerian account, the role of sounds in affording the perceiver with auditory awareness of distal sources is limited to being an audible media through which, or in which, their sources may be heard.

% Hearing the sources of sound is a special, if privileged, case of audition. We may hear a sound and struggle to identify its source, say. So explaining the presentation in audition of the source of a sound in terms of the principle of sympathy is not yet to offer a general explanation of auditory presentation in terms of sympathy. If however a sound that we hear is segmented from all that we hear by identifying, in audition, its source, then the presentation of sources in audition will determine the sounds that we hear, and so sympathy is operative in auditory presentation quite generally. Heidegger was right to emphasize the difficulty in listening away from the sources of sound. If one can only hear a sound by being explicitly aware of its source, then listening away from the source would involve the loss at least of the ability to segment that sound from all that was heard.

% section the_heideggerian_alternative (end)

\section{The Function of Audition} % (fold)
\label{sec:the_function_of_audition}


Begin with two claims recently defended by Nudds:
\begin{enumerate}[(1)]
	\item The function of auditory perception is to afford the perceiver with awareness of the distal sources of sound;
	\item In hearing a sound in a complex sonic environment with multiple active sources of sound, the sound the perceiver hears is segmented from all that they hear due, in part, to their auditory system identifying its source.
\end{enumerate}

Concerning the first claim Nudds writes:
\begin{quote}
	It is uncontroversial to suggest that auditory perception tells us about the sources of sounds as well as about sounds. The suggestion that I am going to develop is that the function of auditory perception is to tell us about the sources of sounds---that perceiving the sources of sounds is what auditory perception is for and that what sounds we hear we hear as a consequence of the particular way auditory perception functions to tell us about the sources of sounds. \citep[284]{Nudds:2010aa}
\end{quote}
The function of auditory perception is to afford the perceiver with awareness of the distal sources of sound. This is a teleological claim. The end of auditory perception, that for the sake of which perceivers are equipped with audition, is the presentation, in audition, of distal events in the natural environment. It is also an explanatory claim. The operation of audition adequate to its function constitutes an explanatorily relevant kind. It is also objective. The operation of audition adequate to its function constitutes an explanatorily relevant kind independently of whether anyone accepts that it does. Moreover, this objective, teleological, explanatory claim, seems naturalistically acceptable. It is, at any rate, overwhelmingly plausible to suppose that an animal's ability to hear distal events in the natural environment contributes to its fitness. And if that is right, that the function of audition is to present distal events in the natural environment is plausibly determined by evolutionary pressures.

Nudds' claim about the function of audition generates a tension within the Peripatetic framework. Consider the following two claims about the proper sensibles:
\begin{enumerate}[(1)]
	\item Proper sensibles are perceptible to one sensory modality alone (for example, one can see colors, but not hear, smell, taste, or touch them)
	\item Proper sensibles are the final cause of perception (for example, sight is for the sake of seeing colors in the light and the luminous in the dark)
\end{enumerate}
The difficulty is that, at least in the case of audition, these two claims cannot be true together.

Consider the second claim first, that the proper sensibles are the final cause of perception. The proper object of sight is the visible (\emph{De Anima} 2 7 418\( ^{b} \)27) and there are two kinds of \emph{visibilia}, color which is visible in light and the luminous, such as bioluminescence or starlight, visible only in the dark (\emph{De Anima} 2 7 419\( ^{a} \)1--7; the nice example of starlight is due to Philoponus  \emph{On \emph{De Anima}} 347 11). If sight is for the sake of seeing colors in the light and the luminous in the dark (\emph{Metaphysica} \( \Theta \) 8 1050\( ^{a} \)10), then is audition for the sake of hearing sounds? Nudds denies this, claiming, instead, that the function of audition is to afford the perceiver with awareness of distal events in the natural environment. Suppose, then, that audition is for the sake of hearing distal sources. Arguably it is that in which audition's selective advantage lies. Hearing sounds would be incidental to audition, so conceived, at least relative to its end, even if one can only ever hear sources through, or in, the sounds they generate. The difficulty is that the final cause, the distal sources, are perceptible to more than one sense alone. Thus one might hear one of London's feral parrots calling as one sees that parrot calling. So there is no one thing that is audible yet perceptible to no other sensory modality and that for the sake of which we possess audition. (1) and (2) are generalizations that fail for the case of audition if we accept Nudds' claim. Perhaps sounds are audible and perceptible to one sense alone and so would make (1) true but (2) would fail---audition is not for the sake of hearing sounds but their sources. The audible sources of sound would make (2) true but (1) would fail---audible sources may be available to more than one sense.

That (1) and (2) fail to be jointly true of audition signals the breakdown of the guiding explanatory framework of \emph{De Anima} 2:
\begin{quote}
	It is necessary for the student of these forms of soul first to find a definition of each, expressive of what it is, and then to investigate its derivative properties, \&c. But if we are to express what each is, viz. what the thinking power is, or the perceptive, or the nutritive, we must go farther back and first give an account of thinking or perceiving; for activities and actions are prior in definition to potentialities. If so, and if, still prior to them, we should have reflected on their correlative objects, then for the same reason we must first determine about them, i.e. about food and the objects of perception and thought. (Aristotle, \emph{De Anima} 2 4 415\( ^{a} \)14--22\index{De Anima@\emph{De Anima}}; Smith in \citealt[26]{Barnes:1984uq})
\end{quote}
Aristotle's explanatory strategy has two parts. 

First, Aristotle proposes to explain perceptual capacities in terms of what they are the capacity for, perceiving.  Specifically, perceptual activity is prior in account to the potential for such activity, the relevant perceptual capacity. Possessing a capacity is a way for things to be, and what it is to be that way depends upon what it is to be its exercise. So possessing audition is a way at least for animals to be, and what it is to be that way depends upon what it is to hear. Thus, if capacities are powers or potentialities, as Aristotle conceives of them, then they ontologically depend upon what they are the potential for. (On ontological dependence see \citealt{Fine:1995ls}. On this reading of priority in account see \citealt{Peramatzis:2011aa}. For a contemporary defence of this claim see \citealt{Kalderon:2012fk}.) 

Second, perceptual activities, the exercise of our perceptual capacities, are themselves partly explained in terms of their correlative objects. It will emerge that, at least with respect to perception, Aristotle means, more specifically, proper objects, understood as sensible objects perceptible in themselves and perceptible to that sense alone. So what it is to hear depends upon the presentation, in auditory experience, of the proper object of audition, sound. Crucially, that is consistent with auditory experience presenting more than just sound.

Thinking of perceptual capacities as individuated by that for the sake of which they are a potential for, allows Aristotle to think that there are exercises of our perceptual capacities that are not the presentation of the proper sensibles, notably, when they are the presentation of common or incidental sensibles. And more besides---the difference between proper objects such as color and sound are perceptible as well. In this way, Aristotle broadens the domain of the perceptible \citep{Sorabji:1971fr,Sorabji:2003fk}. Sight may enable a perceiver to see colors in the light and the luminous in the dark, but it enables the perceiver to see other \emph{visibilia} such as motion, a common sensible. But the presentation of motion in sight is incidental to its operation. Sight is for the sake of seeing colors in the light and the luminous in the dark. But in continuing to understand the presentation of proper sensibles as that for the sake of which a perceiver possesses the relevant perceptual capacity, Aristotle cleaves too closely to the Platonic tradition undone by audition whose function is to afford the perceiver awareness of distal events in the natural environment that are perceptually available to other sensory modalities, such as a storm whistling in the chimney or the call of a feral parrot. One may feel the storm whistling in the chimney and see the parrot calling.

The case of audition undermines the second part of Aristotle's explanatory strategy but not the first. Nothing about audition's function is inconsistent with perceptual activity being prior in account to the capacity for such acitvity, that perceptual capacities are individuated by what they are the potential for, and so ontologically depend upon their proper exercise. Rather, Nudds' claim is a challenge to the second part of Aristotle's explanatory strategy, inherited from Plato if liberalized, that perceptual activities are, in turn, to be explained in terms of the presentation of their proper objects.

% section the_function_of_audition (end)

\section{Sources and the Discrimination of Sound} % (fold)
\label{sec:sources_and_the_discrimination_of_sound}

Audition is for the sake of hearing the sources of sounds, understood as sound-generating events or processes. If audition is for the sake of hearing, not sounds, but their sources, then hearing sound is incidental to audition, relative to its end, even if one can only ever hear sources through, or in, the sounds they generate. Perhaps, in certain contexts, one may even say that one hears a source by hearing its sound, but only in a sense unavailable to the neo-Berkelean (Nudds sometimes writes this way). In Peripatetic terminology, this is an instance of hypothetical necessity (\emph{Physica} 2 9; for useful discussion see \citealt{Charles:1988as}). The necessity is hypothetical since the end of audition, to hear the sources of sounds, is presupposed. Given the end of audition, to hear the distal sources of sounds, it is necessary to hear the sounds that they generate.

Recall, according to the neo-Berkelean, sounds are distinguished from their sources in auditory experience in that only the former are the immediate objects of audition and that we hear the latter by hearing the former. For the neo-Berkelean, the preposition ``by'' is a place holder for the presentative function of sounds, their presenting their sources in presenting themselves in audition. What is this presentative function? One of the lessons we learned from Heidegger was that there is more to the presentative function of sound than the sources we hear being necessarily accompanied, in audition, by their sounds, since sounds may lack this presentative function and remain a necessary accompaniment of the sources that we hear. Moreover, the mediate presentation of sources by the immediate presentation of their sounds is unlike more ordinary cases of perceiving one thing by perceiving another, so in what does this extraordinary case consist? Typically, neo-Berkeleans are no more forthcoming than sense-datum theorists were in giving an account of this presentative function.

According to the neo-Berkelean, sounds are the immediate objects of audition in something like the following sense. Sounds are audible. Moreover, sounds are audible in themselves. Sounds are audible in themselves in the sense that they contain within themselves the power of their own audibility. So one can hear a sound without hearing any other thing. In hearing a sound, auditory experience affords the perceiver with explicit awareness of that sound independently of hearing any other thing. In this sense are they at least among the immediate objects of audition. Though, of course, neo-Berkeleans typically follow Berkeley in maintaining, as well, that sounds are the only immediate objects of audition. Sounds alone have within themselves the power of their own audibility, even if neo-Berkeleans do not go so far as Berkeley in maintaining that sounds are the only objects of audition. \emph{Pace} Berkeley, sources too are audible. However, they are not audible in themselves. They do not contain within themselves the power of their own audibility but are only audible by hearing other objects that are audible in themselves, namely the sounds that they generate. In this sense are they the mediate objects of audition.

So understood, sound could not be the immediate object of audition. Bracket, for the moment, worries about the, as of yet, unexplained presentative function of sound in auditory experience. Focus, instead, on the prior claim that sounds are audible in themselves, that sounds contain within themselves the power of their own audibility with the implication that hearing a sound does not require hearing any other audible object. Nudds' second claim, if true, suffices to establish that sounds are not audible in themselves in the way that the neo-Berkelean requires. Specifically, Nudds claims that in hearing a sound in a complex sonic environment with multiple active sources of sound, the sound the perceiver hears is segmented from all that they hear due, in part, to their auditory system identifying its source. And, as we shall see, that is inconsistent with sounds being audible in themselves. If anything, something like the reverse is true. Sounds are audible, but not audible in themselves, but audible only insofar as one hears the sources that generate them. There is then, I shall suggest, a sense in which sounds are better thought of as audible media through which, or in which, sources may be heard. At the very lest, if sounds were audible media, in the intended sense, they would be audible, but not audible in themselves, but owing their audibility to other things---the sources heard through, or in, the sounds.

Audition, like vision and tactile perception, involves grouping, segmentation, and recognition \citep{Bregman:1990aa}. When, upon the hill in Greenwich Park near the Royal Observatory, I witnessed the Ballardian spectacle of feral parrots traversing the skyscrapers of the City of London, the call of the feral parrots was not all that I heard. I could hear, as well, the trees rustling in the light breeze, the occasional shouts of children playing, people conversing, a bicycle braking, dogs barking. Like most public spaces, Greenwich Park is a complex sonic environment with multiple active sources of sound, and the call of the feral parrots was not all that there was to hear. The patterned disturbance reaching my ears was not solely caused by the parrots calling. And yet I could hear it clearly.

A patterned disturbance, occurring in a given temporal interval, can be analyzed into frequency components, component sine waves of a given frequency and amplitude. When longitudinal pressure waves superimpose, their frequency components additively combine to produce a new complex pressure wave. Given the detected frequency components of the complex pressure wave are not solely caused by the call of the feral parrot, how does my auditory system afford me the capacity to hear the sound of the feral parrot? The auditory system would need to somehow group together the frequency components that constitute the sound of the feral parrot's call.

According to \citet{Nudds:2009sf,Nudds:2010aa}, the auditory system groups frequency components by exploiting clues as to the likely source of the sound. There are a variety of different such clues, and many can be dominated by other clues. 

Some clues are synchronic. That is, sometimes frequency components occurring at a time are related in such a way that it is unlikely that they are the products of distinct sources. For example, the vibration of a material object will determine frequency components of the patterned disturbance that are harmonically related to a fundamental frequency. So there is a tendency for the auditory system to group together frequency components at a time that are harmonically related since it is unlikely that they are produced by distinct sources. 

Some clues are diachronic. That is, sometimes frequency components occurring over time are related in such a way that it is unlikely that they are the products of distinct sources. Thus, for example, the frequency components of a sound produced by a source will have the same onset time, and they will change over time in similar ways. So there is a tendency for the auditory system to group together frequency components that are diachronically related in important ways since it is unlikely that they are produced by distinct sources. 

\citet[74]{Nudds:2009sf} observes that while the clues discussed so far are ``bottom-up'' or stimulus driven groupings, there are, as well ``top-down'' groupings, especially of sequences of frequency components. The idea is that certain frequency components are grouped together because they fit together to form a pattern recognized by the auditory system to likely be produced by a single source. Thus, for example, one might hear a bottle bouncing, as opposed to breaking, and this is likely to be due to such a top-down grouping.

Notice how the clues to grouping together frequency components constituting a heard sound are all based on features of its material source. Thus, for example, the size of an  object will determine the lowest frequency at which it will vibrate. This allows us to hear that one object dropped is larger than another object that is also dropped. How exactly the auditory system extracts information about the material source and what information it extracts from the grouped frequency components is presently not well understood. 

However, exactly, the auditory system performs this feat, the important point is that in a sonically complex environment with multiple active sources of sound, an individual sound is segmented from all that is heard, in part, by identifying its source. If a likely source is not identified by the auditory system, then the frequency components will not be grouped together and the sound will not be segmented from all that is heard. If in hearing the products of multiple active sources of sound, none of the sources are discriminated, all that would be heard is a kind of undifferentiated noise. Hearing the sources of sound lends intelligibility to what is heard. The audible accessibility of the source intelligibly differentiates, in auditory experience, the sound it generates.

At least, then, in ordinary cases of hearing ecological sound, the sound is segmented from all that is heard only by virtue of the auditory system identifying its source. In such cases, sounds are not, in fact, audible in themselves. It is not possible to hear the sound segmented from all that is heard in and of itself quite apart from hearing anything else. In such cases, one hears the sound only insofar as one hears the event or process that generates the sound, only insofar as one hears its audible source. At least in sonically complex environments with multiple active sources of sound, sounds are not audible in themselves, they do not contain within themselves the power of their own audibility, but are only audible insofar as one hears their sources as well. 

The neo-Berkelean shares with Berkeley the conviction that sounds are audible in themselves. They depart from Berkeley in maintaining that we hear, in addition to sounds, their sources. These latter are not audible in themselves but are only audible insofar as we hear their sounds. But the sounds we hear in sonically complex environments are not audible in themselves as the neo-Berkelean conceives of them. Nor, in such circumstances, is there an explicit experience of sound had independently of hearing anything else. So there is no explicit experience of sound to mediately present its source as the neo-Berkelean requires.

% section sources_and_the_discrimination_of_sound (end)

\section{Sympathy and Auditory Presentation} % (fold)
\label{sec:sympathy_and_auditory_presentation}

Sounds are not audible in themselves. Rather, they are more like audible media. What does it mean to describe sounds as perceptual media? Perceptual media need not be thought of as physical media, the movement of whose local parts, ``the oscillation of certain particles,'' realize the progressive instantiation of a wave form. While the idea of physical media merely answers to the demands of being a causal intermediary, the idea of perceptual media answers to the demands of perceptual accessibility. So consider the following. Just as illumination makes the visible perceptually accessible, sound makes the activities of distal bodies perceptually accessible. Without illumination, the colors of distal bodies remain unseen, without sound, the activities of distal bodies remain unheard. (Absent the ring of Gyges, \emph{Republic} 2 359a--360d, becoming invisible is not possible for us, but becoming inaudible is easy enough---simply stop moving.) One sees through, or in, illuminated media, such as air or water, and thereby perceives the colors of distal bodies arrayed in the natural environment. One hears through, or in, audible media, the sound, and thereby perceives the activities of distal bodies arrayed in the natural environment.

By means of the propagation of light waves, the visible aspects of distal bodies are seen. By means of the propagation, in all directions, of the patterned disturbance through a dense and imperfectly elastic medium, that is, by means of sound, the audible activities of distal bodies are heard.

Sound, like the illuminant, is perceptible. Moreover, sound, like the illuminant, is perceptible in a certain way. Concerning the perception of the illuminant, Hilbert writes:
\begin{quote}
	Do we see how an object is illuminated or do we see the illumination itself? On phenomenological grounds the first option seems better to me. What we see as changing with the illumination is an aspect of the object itself, not the light source or the space surrounding the object. \citep[150--151]{Hilbert:2007qy}
\end{quote}
One sees the character of the illumination by seeing the way objects are illuminated. When viewing a brightly lit pantry, one sees the brightness of the pantry by seeing the brightly lit objects arranged in it. So the illuminant is visible, though not visible in itself, but owes its visibility to the objects that it illuminates. (For a comparison with Aristotle's definition of transparency, \emph{De Anima} \textsc{ii} 7 418\( ^{b} \)4--6, see \citealt[41--42]{Kalderon:2015fr}.)

Like the illuminant, sound is perceptible, though perceptible in a certain way. One hears the character of a sound by hearing the activities of its distal source. (Think of how difficult it is to describe ecological sound without describing audible aspects of its source.) So sound is audible, though not audible in itself, but owes its audibility to the distal source that it discloses.

Bregman describes a game that provides a useful analogy:
\begin{quote}
	The game is this. Your friend digs two narrow channels up from the side of the lake. Each is a few feet long and a few inches wide and they are spaced a few feet apart. Halfway up each one, your friend stretches a handkerchief and fastens it to the side of the channel. As waves reach the side of the lake they travel up the channels and cause the two handkerchiefs to go into motion. You are allowed to look only at the handkerchiefs and from their motions to answer a series of questions: How many boats are there on the lake and where are they? Which is the most powerful one? Which is the closer? Is the wind blowing? Has any large object been dropped suddenly into the lake? Solving this problem seems impossible, but it is a strict analogy to the problem faced by our auditory systems. The lake represents the lake of air that surrounds us. The two channels are our two ear canals, and the handkerchiefs are our ear drums. The only information that the auditory system has available to it, or ever will have, is the vibrations of these two ear drums. Yet it seems able to answer questions very like the ones that were asked by the side of the lake: How many people are talking? Which one is louder, or closer? Is there a machine humming in the background? \citep[5--6]{Bregman:1990aa}.
\end{quote}

One striking aspect of Bregman's analogy is how it presupposes that the function of the auditory system is to afford awareness of distal events and processes in the natural environment. The game is to figure out how such awareness is afforded by sensitivity to proximal perturbations in the surrounding medium, be it air or water. The proximal perturbations in the surrounding medium, the patterned disturbances impinging upon the perceiver, while there to be sensed, considered in and of themselves, are relatively unimportant features of the natural environment. Indeed, in Bregman's proposed game, he takes such sensitivity for granted. What is important is not the sensitivity to proximal perturbations in the medium, but sensitivity to the information they carry about distal events and processes in the natural environment, for it is in virtue of this latter sensitivity that the auditory system affords the perceiver with auditory awareness of the distal environment.

The ear channels the longitudinal pressure waves into the ear canal where they come into conflict with the tympanic membrane. The potential pattern of activation of local receptors constitutes the sensitivity to such proximal perturbations. As Bregman's analogy brings out, the task of the auditory system is to somehow extract information about the distal sources of the proximal perturbations. Nevertheless, quite apart from this central role, the end of audition, to afford the perceiver with auditory awareness of the distal environment, the proximal perturbations, the force of the patterned disturbance coming into conflict with the tympanic membrane, are there to be sensed, even if, as Heidegger insists, they are rarely if at all attended to, in familiar every day instances of hearing ecological sound.

Suppose, hypothetically, there could be an auditory experience that arose from this sensitivity to proximal perturbations, quite apart from what information they could provide about distal sources, so that the experience was confined to only what was proximately impinging upon the perceiver. Perhaps auditory experience upon first regaining consciousness may approximate what we are presently supposing. What would be experienced would be a kind of undifferentiated noise. Even in the case where the patterned disturbance was produced by a single source, there is a marked difference in audible intelligibility between hearing the sound as produced by that audible source and hearing the sound and being unaware of its source. Upon first coming to consciousness the perceiver hears a sound but cannot make out its source. As things come into focus for the perceiver, they come to hear the source of the sound. In hearing the sound, there is a marked increase in the intelligibility of what is heard. What is heard is no longer mere noise but the storm whistling in the chimney or the call of a feral parrot, say.

But even hearing a noise, like the case of grasping or enclosure, understood as a mode of haptic perception, would involve the presentation of something extrasomatic. The longitudinal pressure waves may be impinging upon the perceiver, but they are extrasomatic for all that. This hypothetical limited auditory experience is no mere auditory sensation but a mode of auditory perception, the presentation, in auditory experience, of an extrasomatic event or process. And like the case of grasping or enclosure, understood as a mode of haptic perception, it is sympathy that makes the presentation in conscious experience of the extrasomatic possible. The force of the propagation of the patterned disturbance comes into conflict with the force of the tympanic membrane and this gives rise to the hypothetical limited auditory experience. Moreover, this limited auditory experience is no mere auditory sensation, a conscious modification of the perceiving subject brought about by impingement from without, but the perception of extrasomatic events or processes. Even supposing that the impingement of proximal perturbations occasioned in the perceiver intensive auditory sensation, such sensation is only the presentation of this noise insofar as it is experienced as a sympathetic response to an extrasomatic event. Sympathy is what makes for this difference. The hypothetical limited experience that auditory sensitivity to proximal perturbations gives rise to involves the sympathetic presentation of those perturbations. 

One important difference between this hypothetical limited form of auditory experience and grasping or enclosure, understood as a mode of haptic perception, is the relative passivity of the former compared to the latter. In the conflict between the force of the propagation of the sound, the dynamic principle of unity of the sound event, and the countervailing force determined by the tension and elasticity of the tympanic membrane, the former acts upon the latter and the latter merely resists the former insofar as it can (there are limits, of course, one can blow an ear drum). In grasping or enclosure, understood as a mode of haptic perception, by contrast, the hand, unlike the tympanic membrane, is active. Indeed, it is the active wax of haptic perception.

The limited auditory experience is hypothetical, even if actual auditory experiences, such as those undergone when coming to consciousness, may approximate it. However, when we consider more ordinary cases of auditory perception, such as hearing the storm whistling in the chimney, the three motored plane, the Mercedes in immediate distinction from the Volkswagen, not only does sympathy play an expanded role in the presentation, in audition, of the activities of distal bodies, but, moreover, this sympathetic presentation is made possible by the perceiver listening out for distal events or processes, thus reinstating the analogy with haptic perception.

The perceiver hears the distal source in the sound that it generates. Moreover the perceiver hears in conformity with the sound that the distal source generates. And it is the principle of sympathy that governs the disclosure, over time, of the distal sound-generating events and processes.

The perceiver hears the distal source in the sound that it generates. On one construal of Bregman's analogy, perhaps not the only one, hearing the sources of sound through, or in, the sounds they generate is not unlike the quasi-visual perception induced by Bach-y-Rita's tactile-visual substitution system (for an overview see \citealt{Bach-y-Rita:2002jk}). Bach-y-Rita's tactile-visual substitution system involved a head mounted camera wired to electrodes attached to the perceiver's body. The idea was to map the visual information captured by the camera onto a pattern of tactile activations. Perceivers that were able to control the camera by ``looking'' around could, within a day, make quasi-visual reports about the number, size, and relative distance of objects arrayed in their environments. At this point, perceivers were explicitly aware, in a quasi-visual mode of awareness, of distal objects in the natural environment and merely implicitly aware of any tactile sensation. In contrast, at the beginning of this procedure, when the apparatus was first mounted and used, the perceiver was only explicitly aware of the electrodes' stimulation. Their experience only came to quasi-visually present distal objects and their spatial properties when they learned to sympathetically respond to the electrodes' stimulation. In so doing, they learn to ``see'' distal objects through, or in, what they feel. Moreover, if sympathy is indeed the principle governing this quasi-visual presentation, then that would explain the pattern of attention described above. If the perceiver, having mastered the tactile-visual substitution system, were to explicitly attend to the electrodes' stimulation, this would erode the sympathetic presentation of the distal objects and their spatial properties. Similarly, in hearing a distal event or process, such as hearing the storm whistling in the chimney, the three motored plane, the Mercedes in immediate distinction from the Volkswagen, the perceiver is explicitly aware of the source and only implicitly aware of the sound it generates. They hear the source through, or in, the sound it generates. Moreover, if sympathy were the principle of the disclosure, in audition, of distal events and processes, this would explain this pattern of attention. If the perceiver could indeed listen away from the source, listen abstractly, and attend only to the sound it generates, this would erode the sympathetic presentation of the distal source. 

The perceiver hears in conformity with the sound that the distal source generates. In hearing distal events and processes in the natural environment, the perceiver is explicitly aware of these sources. However, reflection on perceptual constancy reveals that the phenomenological character of auditory experience is not exhausted by the object of explicit awareness. An implicit awareness of the sound they generate contributes, as well, to the phenomenology of the perceiver's auditory experience. Specifically, an implicit awareness of sound contributes to the way in which the explicit object of the auditory experience is presented therein. So consider approaching a continuous source of sound, such as a waterfall. The waterfall, heard from different distances, sounds different. Heard from afar, the waterfall sounds quieter than it does when heard from nearby. As the perceiver approaches the waterfall, the sound of waterfall increases in volume. But throughout the perceiver's approach, the perceiver heard the constant flowing of the waterfall. The flowing of the waterfall is not experienced as getting louder so much as the perceiver is getting in a better position to hear just how loud the waterfall really is. The flowing of the waterfall, the constant object of explicit auditory awareness is not experienced as changing in the way that it would have to if it were in fact getting louder, only its auditory appearance is changing with a change in auditory perspective. The flowing of the waterfall, the object of explicit awareness and the constant element in the phenomenology of stability and flux, sounds different when heard from different auditory perspectives. Hearing the sound of the waterfall, from a given auditory perspective, may be implicit, it may be recessive and in the background, so that it does not compete for attentive resources directed towards the flowing of the waterfall, but it contributes to the conscious character of the perceiver's auditory experience by being the way in which the distal process is presented in that experience. The auditory disclosure of a distal source just is hearing that source in the sound that it generates and hearing in conformity with that sound. And that just is the exercise of a sympathetic capacity in audition.

In sympathetically presenting the distal sources of sound, auditory experience is constitutively shaped by the distal events and processes it discloses. The conscious character of hearing a watch ticking is constituted, in part, by the audible ticking of the watch. And the conscious character of hearing the call of a feral parrot is constituted, in part, by the feral parrot's call. What it is like for the perceiver to hear the ticking of the watch depends upon and derives from what the ticking of the watch is like---how loud it is, its distinctive timbre, what the mechanism sounds like. And what it is like for the perceiver to hear the call of the feral parrot depends upon and derives from what the call of the parrot is like---how loud it is, its distinctive timbre, its sharpness and urgency. Moreover, what it is like for the perceiver to hear these events and processes depends, as well, upon the perceiver's perspective. There are better and worse perspectives, even if better and worse is said of in many ways. Auditory experience formally assimilates to its object, relative the perceiver's partial perspective, as a consequence of being constitutively shaped by that object as presented to that perspective, a constitutive shaping made possible by the sympathetic presentation of that object in auditory experience. Constitutive shaping of auditory experience by its object is a ``communion'' with that object---in undergoing that experience the perceiver is united, in a way, with the object of their perception. Moreover, as with Plotinus (chapter~\ref{sec:sympathy_as_the_principle_of_haptic_presentation}), this unity explains in part, the similarity between the auditory experience and its audible object. The formal assimilation of auditory perception to its object, at least relative to the perceiver's auditory perspective, is the effect of constitutive shaping, and thus its conscious character depends upon and derives from, at least in part, the audible character of the object heard.

Recall, we are following Plotinus in taking the unity of auditory presentation to be explanatorily prior to the operation of sympathy (chapter~\ref{sec:sympathy_as_the_principle_of_haptic_presentation}). Auditory presentation of distal sources is not being constructed from elements and principles understood independently of their auditory presentation, rather the unity of the perceiver and the distal events and processes is presupposed, and sympathy merely analytically explicates the intelligible structure of this presupposed unity. Not only does sympathy only operate within a unity, but that unity is reducible to no other thing.

Auditory presentation is an irreducible unity. If sensory presentation is a distinctive kind of unity, a ``communion'' with its object, then auditory presentation is more distinctive still. Insofar as auditory presentation, like haptic presentation, is governed by the principle of sympathy, it is a mode of being with. Hearing the call of the feral parrot is a way of being with that bird at least insofar as it is engaged in audible activity. Whereas haptic presentation is corporeal, it is a way for a conscious animate body to be with another body, auditory presentation, while involving a conscious animate body, is not completely corporeal, since it involves a conscious animate body, the perceiver, being with an event or process (which may or may not involve bodies as participants). Distal bodies are never present in auditory experience \emph{qua} bodies, but only as audible participants of an unfolding audible event or process. Auditory presentation, like haptic presentation, is a kind of disclosure with duration. However, whereas the objects of audition are not wholly present at any given moment, the objects of haptic perception may be, as when one feels relatively static features of bodies such as their texture or temperature. \emph{Audibilia}, on the other hand, are essentially dynamic entities not wholly present at any given moment. They unfold through the temporal interval of their sounding. If heard, they are disclosed, in audition, over time. In hearing something, we listen along with it.

% section sympathy_and_auditory_presentation (end)

\section{Listening} % (fold)
\label{sec:listening}

Let us return to the Protagorean model (chapter~\ref{sec:haptic_perception}) and the challenges it faces in applying to audition (chapter~\ref{sec:moving_forward}). According to the Protagorean model, perception is the joint upshot of forces in conflict. Grasping or enclosure, understood as a mode of haptic perception, is itself naturally understood on this model. On the one hand, there is the force of the activity of the grasping hand. On the other hand, there are the self-maintaining forces of the rigid, solid body. Making an effort to more precisely mold the hand to the body's contours and the resistance of the self-maintaining forces that determine that body's rigidity and solidity together give rise to an experience of that body's overall shape and volume. 

Can auditory perception be understood on the Protagorean model, as the joint upshot of forces in conflict? Consider, again, the hypothetical auditory experience whose content is limited only to the proximal perturbations in the local medium. On the one hand, there is the force of the tympanic membrane, determined by its tension and elasticity. On the other hand, there is the force with which the patterned disturbance propagates in the dense and elastic medium, the dynamic principle of unity of the sound event. The force of the patterned disturbance coming into conflict with the force of the tympanic membrane gives rise to the perception of an extrasomatic event, the proximal perturbations, the force of the sound event impinging upon the perceiver, even if it is only heard as a mere noise.

In the previous section, we noted a crucial disanalogy with the case of grasping or enclosure, understood as a mode of haptic perception. Specifically, the present application of the Protagorean model to the hypothetical limited auditory experience is passive. The force of the growth of the sound event acts upon the tympanic membrane occasioning a pattern of local activations that give rise to, given subsequent processing, an auditory perception. In contrast, the hand, in grasping or enclosure, understood as a mode of haptic perception, is not merely acted upon by the object grasped but actively grasps that object. The hand actively assimilates to its object. The hand is, in this way, the active wax of haptic perception.

I believe that there is a conception of listening more active than the passive power of the tympanic membrane to receive stimulation, and that this more active conception of listening will partially restore the analogy with grasping. But before I explain further allow me to briefly discuss an important, historically salient variant of the present difficulty.

``In order to hear well,'' Maine de Biran observes, ``it is necessary to \emph{listen}'' (\emph{Influence de l'habitude sur la faculté de penser}; \citealt[63--4]{Boehm:1929aa}). Listening---like grasping, feeling weighing, and looking---is active. It is something that the perceiver does. Grasping, feeling, weighing, looking, and listening are not voluntary intentional movements, though they may involve these. Rather, each are a kind of perceptual stance, sustained by a characteristic activity, where the perceiver opens themselves up, in a directed manner, to experiencing different aspects of the natural environment.

Listening, for Maine de Biran, is ``the putting into action the muscles destined to communicate different degrees of tension to the membrane of the tympanum, etc'' \citep[64]{Boehm:1929aa}. Maine de Biran is engaging in speculative anatomy, here. Just as muscles attached to the eye can expand or contract the pupil in order to better see, Maine de Biran's thought is that muscles attached to the tympanic membrane tighten or loosen it in order to better hear. However, the efforts involved in such motor activity are ``imperceptible'' and ``do not manifest themselves at all as expressions of the will'' \citep[64]{Boehm:1929aa}. Compare the distinction that \citet[]{Smith:2002sa} draws between saccadic eye movement and deliberately moving the eye in its socket. The former is involuntary and so does not manifest itself as the expression of the will the way the latter does. How, then, is the analogy with haptic perception sustained? 

Main de Biran PROVIDES a providential response:
\begin{quote}
	But nature herself has taken care to supplement these faults; she has restored equilibrium by associating in the most intimate way her passive impressions with the activity of an organ essentially motor. (\emph{Influence de l'habitude sur la faculté de penser}; \citealt[63--4]{Boehm:1929aa})
\end{quote}
The ``organ essentially motor'' is, more specifically the vocal organ. Through the effects of natural sympathy rendered insensible by habit, the vocal organ engages in a kind of subvocalized echoing of heard sound. So sounds impinging upon the perceiver cause passive auditory impressions, and the vocal organ ``repeats them, imitates them, turns them back, if one might say so, towards their source, and afterwards makes these fleeting modifications enter the sphere of the individual's activity, establishes them and incorporates them there'' \citep[64]{Boehm:1929aa}. It is the habitual sympathetic activity of the vocal organ, echoing the auditory character of the passive impression caused by the sound, that presents the sound to consciousness. Moreover, the presentation in auditory experience of sound is described as a kind of incorporation, an image at the center of the semantic field of metaphors loosely organized as modes of assimilation (chapter~\ref{sec:grasping_and_the_dawn_of_understanding}). With this active echoing of passive impressions, Maine de Biran claims to restore the analogy with haptic touch.

It is easy to be suspicious of this providential natural supplement (for criticism see \citealt[chapter 7]{Derrida:2005aa}). However, if confined to the special case of speech perception, the natural supplement is more plausible. Indeed, Maine de Biran's principle example is following along under our breath in hearing singing or speaking. The natural supplement is charitably understood as an overgeneralization from the special case of speech perception, on a particular understanding of speech perception, where our capacity to perceive speech draws upon our capacity to produce speech (compare \citealt[chapter 2]{Bergson:1912pi}, \citealt{Liberman:1985ty}, \citealt{Mole:2009hl}). 

What motivated the Biranian doubling of passive impression with vocal activity was the thought that to fully restore the analogy with haptic touch an actual activity must be found. The temptation to overgeneralize from the special case of speech perception may be avoided if we relax this demand. Perhaps, listening, the activity of listening out for something, is a perceptual stance sustained by the potential for such activity, activity that will make for a better or worse perspective from which the audible object may be heard. Thus, for example, our hominid ancestor, in hunting tapir, may pause to listen out for movement in the bush. In listening out for movement, our hominid ancestor may be prepared to turn in the direction of the heard movement, to better attend, in audition, to such movement. They are prepared to turn, and listen, and hear. While many forms of listening involve actual activity on the part of the perceiver, listening, understood as a perceptual stance, may be sustained by the potential for such activity.

This last claim needs qualification. It is not the bare possibility of the perceiver responding in ways that will alter their auditory perspective to better hear what there is to be heard that sustains the relevant perceptual stance. In one sense, that much is possible even should the perceiver be unconscious. In another sense, being unconscious, it is not possible for them to respond in an appropriate way to auditory stimuli. Not only must the perceiver be conscious, in order for it to be possible for them to engage in the relevant activity, but more stringently still, the perceptual stance must be sustained by the preparedness to act in these ways should the circumstances warrant it, given the practical ends in play in those circumstances. That the relevant sense of potential activity involves the preparedness to act is more stringent still since the preparedness to act would involve changes to the perceiver not present when a perceiver, though conscious, is not so prepared.

With this qualification in place, the analogy with haptic touch is partially, if not fully, restored. Grasping or enclosure involves the activity of the hand, while listening merely requires a preparedness to act in auditorily relevant ways. While we have found a role for activity in hearing aspects of the distal environment, listening merely requires the potential for such activity, understood as a preparedness to engage in such activity. Audition remains not as active as the exemplar, grasping, but it is not merely passive the way the registering of movement by the tympanic membrane is. Listening requires the perceiver's vigilance. But the perceiver's auditory vigilance over the distal environment, their being prepared to act in auditorily relevant ways to bring aspects of the distal environment into earshot, remains a stance actively sustained by the perceiver.

In the traditional, post-Aristotelian vocabulary, the distinction between listening and grasping can be described in terms of first and second actuality. The distinction is traditionally introduced in terms of Aristotle's discussion of knowledge in \emph{De Anima} 2 3 (417\( ^{a} \)22--417\( ^{b} \)1). Thus an educable person may be ignorant of some point of grammar. But since they are educable, learning that point of grammar is not beyond their ken, and so they may be said to, in this sense, potentially know that point of grammar. Suppose the ignorant if educable person comes to learn it. In learning the relevant point of grammar, they come to actually know it. But, Aristotle observes, the knowledge of the now learned person is itself a kind of potentiality. It is the capacity to apply that knowledge in a reasonable manner given the practical circumstances. Thus the learned person might reasonably apply their knowledge of grammar in interpreting the speech of another. In reasonably applying their knowledge in the given circumstance the learned person actualizes their knowledge. Learning is the actualization of the educable person's capacity for knowledge. In the traditional post-Aristotelian vocabulary, it is the first actuality. But since what is learned, knowledge of the relevant point of grammar, is itself a kind of potentiality---it is the capacity to apply that knowledge in a reasonable manner given the practical circumstances---its exercise is itself a kind of actualization. It is the second actuality. Grasping requires the second actuality of the hand's activity whereas as listening merely requires a first actuality, or equivalently, a second potentiality---the capacity to act in auditorily relevant ways. Nevertheless, this second potentiality, required by listening, involves the preparedness to act, a kind of perceptual vigilance, which itself requires activity on the part of the perceiver to sustain.

I turn, and listen, and hear the call of the feral parrot. What I hear is the audible activity of a distal body, the animate body of the feral bird. Audition affords me explicit awareness of the parrot's call. I hear how loud it is, its distinctive timbre, and its sharpness and urgency. I hear the parrot's call through, or in, the sound it makes. The parrot's calling generates a patterned disturbance that propagates, in every direction, through the dense and imperfectly elastic air. It is through, or in, this audible media, the sound it makes, that the call of the feral parrot is heard. In turning, and listening, and hearing, I alter my auditory perspective on the natural environment to bring an aspect of that environment, the audible activity of the feral parrot, into earshot. Turning, and listening, and hearing---actively changing my auditory perspective on the natural environment---is itself a sympathetic response to what is heard. Changing my auditory perspective to increase the acuity with which the feral parrot is heard is to sympathetically respond to the call of the feral parrot. Preparedness to act in certain ways so that the impingement of the force of the propagation of the patterned disturbance, the dynamic principle of unity of the sound event, carries information about its distal source, sensitivity to which constitutes, in propitious circumstances, explicit auditory awareness of that source, is what makes possible the sympathetic presentation, in auditory experience, of the source of the sound. The power to receive auditory stimulation from proximal perturbations may be purely passive, but it is the perceiver's perceptual vigilance, their preparedness to alter the circumstances in which such stimulation is received with the end of hearing its distal source, that makes for the sympathetic presentation, in audition, of the source of the sound.

% section listening (end)

% chapter sources_of_sound (end)
