%!TEX TS-program = xelatex 
%!TEX TS-options = -synctex=1 -output-driver="xdvipdfmx -q -E"
%!TEX encoding = UTF-8 Unicode
%
%  extramission
%
%  Created by Mark Eli Kalderon on 2015-10-27.
%  Copyright (c) 2015. All rights reserved.
%

\documentclass[12pt]{article} 

% Definitions
% \newcommand{\change}{\textcolor{blue}{\textbf{CHANGE SLIDE}}}
\newcommand\mykeywords{vision, extramission}
\newcommand\myauthor{Mark Eli Kalderon}

% Packages
\usepackage{geometry} \geometry{a4paper} 
\usepackage{url}
% \usepackage{txfonts}
\usepackage{color}
% \usepackage{enumerate}
\definecolor{gray}{rgb}{0.459,0.438,0.471}
\usepackage{setspace}
% \doublespace % Uncomment for doublespacing if necessary
% \usepackage{epigraph} % optional

% XeTeX
\usepackage[cm-default]{fontspec}
\usepackage{xltxtra,xunicode}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\setmainfont{Hoefler Text}

% Bibliography
\usepackage[round]{natbib}

% Title Information
\title{The Truth in Extramission}
\author{\myauthor} 
\date{} % Leave blank for no date, comment out for most recent date

% PDF Stuff
\usepackage[plainpages=false, pdfpagelabels, bookmarksnumbered, backref, pdftitle={Form Without Matter}, pagebackref, pdfauthor={\myauthor}, pdfkeywords={\mykeywords}, xetex, colorlinks=true, citecolor=gray, linkcolor=gray, urlcolor=gray]{hyperref} 

%%% BEGIN DOCUMENT
\begin{document}

% Title Page
\maketitle

% Layout Settings
\setlength{\parindent}{1em}

% Main Content

\section{The Persistence of Extramission} % (fold)
\label{sec:the_persistence_of_extramission}

\citet[48]{Piaget:1929dp} observed that there is a the tendency for children to understand vision in terms of an active, outward influence of the eyes. This tendency was manifest in reports of looks mixing and in ``a confusion between vision and light''. Concerning the latter Piaget reports:
\begin{quotation}

	Pat (10) stated that a box makes a shadow ``\emph{because the clouds} (Pat believes it to be the clouds which give light when there is no sun) \emph{can't pass through it}'' (\emph{i.e.} because the light cannot pass through the box).
	
	But immediately after Pat said of a portfolio that it made a shadow ``\emph{because the clouds can't see that side}.---Are to see and to give light the same thing?---\emph{Yes}.---Tell me the things which give light?---\emph{The sun, the moon, the stars, the clouds and God}.---Can you give light?---\emph{No} \ldots\ \emph{Yes}.---How?---\emph{With the eyes}.---Why?---\emph{Because if you hadn't eyes you wouldn't see properly}.''
	
	Duc (6 1/2) also stated that the light cannot see through a hand, alike confusing ``seeing'' with ``giving light.''
	
	Sci (6) said that dreams come ``\emph{with the light}.''---``How?---\emph{You are in the street. The lights} (street-lamps) \emph{can see there} \ldots\ \emph{they see on the ground}.'' ``Tell me some things that give light.---\emph{Lights, candles, matches, thunder, fire, cigarettes}.---Do eyes give light or not?---\emph{Yes, they give light}.---Do they give light at night?---\emph{No}?---Why not?---\emph{Because they are shut}.---When they are open do they give light?---\emph{Yes}.---Do they give light like lamps?---\emph{Yes, a little bit}.'' \citep[48]{Piaget:1929dp}
\end{quotation}

And \citet[48--49]{Piaget:1929dp} goes on to compare these reports with Empedocles' lantern analogy \textsc{dk} 31b84. (On the lantern analogy, see \citealt[240--243]{Wright:1981zr}; on Empedocles' theory of vision see \citealt{Sedley:1992uq}, \citealt{Ierodiakonou:2005fk}, \citealt[chapter 1]{Kalderon:2015fr}.) Perhaps the occurence of extramission beliefs early on in our cognitive development explains, in part, the occurence of extramission theories early on in our philosophical history. 

\citet[138]{Winer:1996as}, prompted by Piaget's observations, were ``sur\-pri\-sed---indeed shocked'' by the degree and resilience of belief in extramissive perception. Not only do children hold extramission beliefs but so do adults, though such beliefs tend to decline during adulthood. To the simple question that required a ``yes'' or ``no'' response:
\begin{quote}
	When we look at someone or something, does anything such as rays, waves, or energy go out of our eyes?
\end{quote}
49\% of the first graders, 70\% of the third graders, 51\% of the fifth graders, and 33\% of the college students affirmed extramission. Moreover, these extramission beliefs proved ``highly resistant to experimental intervention designed to alter them'' \citep[138]{Winer:1996as}. 

\citet{Winer:1996as} augmented their use of verbal questions with graphic displays:
\begin{quote}
	The computer graphics portrayed various interpretations of the process of vision by displaying one or more renditions of a person looking at a rectangle, with visual input and output depicted by lines that appeared to move between the person's eye and the rectangle. Thus, in one graphic, lines, presumably representing rays, appeared to move inward from the rectangle to the eye of the figure on the screen, demonstrating the process of intromission. In another graphic, lines appeared to move outward from the eye toward the rectangle, demonstrating pure extramission. \citep[139]{Winer:1996as}
\end{quote}
They did so for two reasons. First, the graphic displays were used to filter out any misinterpretations that might have been suggested by the verbal questions. Second, they predicted that, given a hypothesized source of extramission belief, exposure to the graphic displays would increase the affirmation of extramission.

What is this hypothesized source? \citet{Winer:1996as} hypothesize that both the tendency for extramission beliefs to persist into adulthood and their resistance to experimental intervention is partly explained by a phenomenological truth enshrined in extramission models:
\begin{quote}
	We assume that core aspects of the phenomenology of vision underlie extramission interpretations. Consider one phenomenologically salient aspect of vision, namely, its orientational or outer-directed quality. When people see, they are generally oriented toward an external visual referent, that is, they direct their eyes and attention to an object in order to see it. In fact, this quality of vision is reflected in language. People talk about ``looking at'' things, and English has expressions such as ``looking out of a window'' and ``looking out of binoculars.'' Even notions such as ``piercing glances'' and ``cutting looks'' suggest and outer directionality \ldots\ \citep[140]{Winer:1996as}
\end{quote}
On this basis, they predicted an increase in the affirmation of extramission because of the way that the graphics ``present representations that are suggestive of the orientational aspects of vision.'' And subsequent studies confirmed this.

It is unclear, at least to me, what to make of this increased affirmation of extramission in response to the use of graphic displays. The displays do not unambiguously represent the intended interpretations of the process of vision. Specifically, they do not unambiguously represent lines of causal influence. As \citet{Winer:1996as} observe, they are at least suggestive of the orientational aspects of vision. But given the iconic nature of the pictorial representation, moving lines might represent lines of causal influence, but they might just as easily represent lines of sight. Perhaps, the increased affirmation of extramission is less an expression of belief in extramission than an expression of the active, outer-directed phenomenology of vision. Perhaps, the affirmation of extramission involved belief in, not a scientific misconception, but a phenomenological truth misleading expressed. \citet{Winer:1996as} claim to control for this, but whether they did so successfully is difficult to independently assess.

Whatever the genuine extent of extramission belief, it is the phenomenological diagnosis for it that we shall focus upon. In the next section, we shall examine the active, outer-directed phenomenology of vision that \citet{Winer:1996as} take to underly belief in extramission.

% section the_persistence_of_extramission (end)

\section{The Truth in Extramission} % (fold)
\label{sec:the_truth_in_extramission}

Merleau-Ponty provides a description of the active, outer-directed phenomenology of vision that would make talk of extramission apt:
\begin{quote}
	If I adhere to what immediate consciousness tells me, the desk which I see in front of me and on which I am writing, the room in which I am and whose walls enclose me beyond the sensible field, the garden, the street, the city and, finally, the whole of my spatial horizon do not appear to me to be causes of the perception which I have of them, causes which would impress their mark on me and produce an image of themselves by a transitive action. It seems to me rather that my perception is like a beam of light which reveals the objects there where they are and manifests their presence, latent until then. Whether I myself perceive or consider another subject perceiving, it seems to me that the gaze ``is posed'' on objects and reaches them from a distance---as is well expressed by the use of the Latin \emph{lumina} for designating the gaze. \citep[185]{Merleau-Ponty:1967fj}
\end{quote}
Merleau-Ponty is not endorsing the extramission theory as a causal model of perception. He is not denying that the object of perception is the ultimate efficient cause of that perception. Rather, in seeing the desk before him, Merleau-Ponty claims only that his experience does not present itself as the exercise of a passive power, a sensory impression caused in him by the mediate causal action of the distal object. There may be an active element to outwardly attending, in vision, to distal aspects of the natural environment, and this may be phenomenologically vivid, but that is consistent with the object of visual perception being among its causal antecedents. ``My present experience of this desk is not complete, \ldots\ it shows me only some of its aspects'' \citep[186]{Merleau-Ponty:1967fj}. Merleau-Ponty's experience may be incomplete in that it reveals only some aspects of its object, but once we allow that perception is partial in this way, it is at least open that experience is incomplete, as well, in that it only manifests some aspects of its nature. The active, outer-directed nature of vision may be phenomenologically vivid, but vision may still require that the distal object mediately act upon the perceiver.

Nor is Merleau-Ponty claiming that it appears from within that seeing involves the emission of a fiery effluence akin to light. Rather Merleau-Ponty is pressing an analogy. He is describing what visual experience, from within, is like. And not only from within but from without as well. The analogy holds not only when Merleau-Ponty considers his own experience but also when he considers the experience of another perceiving subject. Consider another's piercing glance or cutting look \citep[140]{Winer:1996as}. Piaget's \citeyearpar[48]{Piaget:1929dp} reports of looks mixing are cases where the analogy would hold, as well, of other perceiving subjects. Merleau-Ponty, then, is describing that aspect of our visual phenomenology, considered from within and without, that \citet{Winer:1996as} claim to underlie extramission beliefs. 

More explicitly, the awareness afforded by visual experience is like a beam of light that manifests the latent presence of its object. Vision, like illumination, has direction. Light is emitted outward from its source upon the scene that it illuminates.  Vision too is outer directed. In seeing, the perceiver looks out upon the scene before them. Not only do vision and light have direction but they are both rectilinear as well. Moreover, just as illumination manifests the latent visibility of an object, seeing an illuminated object manifests its latent presence to the perceiver revealing it to be where it is. The explicit awareness of the natural environment afforded by visual experience is akin to light not only in its rectilinear directionality and its power to manifest latent presence, but in the manner in which it discloses distal aspects of that environment. Just as a beam of light may ``pose'' on an object that it illuminates and that it reaches from a distance, the perceiver's gaze may ``pose'' on the object that it presents and that it reaches from a distance. The imagery here not only emphasizes that vision is a kind of perception at a distance but invokes, as well, an active, outward extension.

Accepting the aptness of the analogy is not tantamount to accepting the extramission theory. Consider a similar analogy of Peter John Olivi's (on Olivi on perception see \citealt[3--26, 39--54]{Tachau:1988aa}, \citealt[215--224]{Spruit:1994qq}, \citealt[121--124, 130--134, 168--181]{Pasnau:1997aa}, \citealt[part 1]{Toivanen:2009zf}, \citealt{Silva:2010zh}, \citealt[part 2]{Toivanen:2013ul}):
\begin{quote}
	an object, to the extent that the gaze (\emph{aspectus}) and the act of a power are terminated at it, co-operates in their specific production [\ldots] Namely, the cognitive act---and the gaze---is fixed (\emph{figitur}) to the object and it absorbs the object intentionally to itself. This is why a cognitive act is called the apprehension of, and the apprehensive extension to, the object. In this extension and absorption the act becomes intimately conformed and assimilated into the object. The object presents itself or appears as being present to the cognitive gaze, and the object is a kind of representation of itself by an act which is assimilated to it. As an actual illumination of a spherical or quadrangular vase becomes spherical or quadrangular only because the light source generates the illumination in conformity with the figure of the object which receives and confines it; so also, because a cognitive force generates a cognitive act with a certain formative absorption of the act towards the object, and with a certain signet-like and inward (\emph{sigillari et viscerali}) extension of the object, therefore---because it is generated thus---the act becomes a similitude and signet-like expression of the object. (Olivi, \emph{Questiones in secundum librum Sententiarum} q. 72 35--36; \citealt[146--147]{Toivanen:2013ul})
\end{quote}
The passage is complex and is replete with suggestive detail. But to begin with, focus on the analogy with illumination.

Despite his play with neo-Platonic imagery, no doubt an Augustinian heritage \citep[198]{Kent:1984zm}, Olivi is not endorsing an extramission theory of perception. Olivi explicitly denies that extension involves any real emission (\emph{Questiones in secundum librum Sententiarum} q. 58 ad 14.8). Perceptual apprehension may be a form of apprehensive extension to its object, but this apprehensive extension is not corporeal. Though likened to illumination directed upon the object it illuminates, the perceptual act, the apprehensive extension by which that act assimilates to its object, does not consist in, or otherwise involve, the emission of a fiery substance, no matter how rarified. Nor does the apprehensive extension involve the emission of any spiritual matter. Perception is a simple, spiritual act that takes place in the soul, and the soul can only be in the body, at least when alive. So no part of the soul is substantially located where the perceived object is as would be the case if extramission were true. Likening the seeing of an object to light directed upon an object that it illuminates, by itself, carries with it no commitment to the metaphysics of extramission. Rather, Olivi, like Merleau-Ponty after him, is emphasizing the active, outer-directed nature of vision.

Moreover, like Merleau-Ponty, Olivi is presenting a conception of perception that contrasts with a mere passive reception of sensible form. According to Olivi, however, the object of perception is not an efficient cause of that perception, no matter how mediate. Olivi is moved, in part, by the Augustinian doctrine of the ontological superiority of the soul over the body. The way in which the soul and its powers and acts are superior to the body is inconsistent with a body ever acting upon the soul. So nothing corporeal could activate the perceptual power, the power and its act being simple and spiritual (\emph{Questiones in secundum librum Sententiarum} q. 73 83--4). \citet[263]{Silva:2010zh} observe how Olivi anticipates, here, the Cartesian distinction between \emph{res extensa} and \emph{res cogitans} \citep[see also][46]{Tachau:1988aa}. The powers of the soul, even perceptual powers, are not the passive recipients of external stimuli but are active. Like Merleau-Ponty, Olivi thinks that the active character of our perceptual powers is phenomenologically evident (\citealt[3--26, 39---54]{Tachau:1988aa}, \citealt[236--47]{Pasnau:1997aa}, \citealt[143]{Toivanen:2013ul}). 

While Olivi does not deny that perception presupposes the presence of its object in the natural environment, he does deny that it is, or even among, the efficient causes of perception. In Olivi's technical vocabulary, the object of perception is a terminative cause. It is controversial how to understand Olivi's terminative causes. Are terminative causes a species of final cause, as \citet[192--195]{Kent:1984zm} and \citet{Pasnau:1999kn} maintain? Or are they a kind of cause not classified by the traditional Peripatetic four causes (\emph{Physica} 2 3), as \citet[chapter 6]{Toivanen:2013ul} maintains? While it is difficult to form a clear, positive conception of terminative causes, the negative contrast with efficient causes is clear. The actualization of a perceptual power may require the presence of its object in the natural environment but that object acting upon the power is not required for its actualization. The efficient cause of the perceptual act is the power and not the object of perception. The presence of that object merely cooperates by being the \emph{terminus} of the perceptual act, that which it is directed upon, like light directed upon a spherical vase.

So Olivi maintains that the active, outer-directed phenomenology of vision is inconsistent with the object seen being the efficient cause of that perception. Merleau-Ponty, by contrast, merely claims that the object of perception acting upon the perceiver, however mediately, is not manifest in our experience, not that it is inconsistent with it. Perhaps this more cautious attitude is, in the end, warranted. I do not recommend this more cautious attitude merely as a beneficiary of optical knowledge unavailable to Olivi but on philosophical grounds as well.

To bring this out, first consider an element of the Olivi passage that goes beyond what Merleau-Ponty explicitly describes. Olivi, like Merleau-Ponty, uses the neo-Platonic imagery of illumination to emphasize the active outward extension involved in the visual apprehension of the distal environment, and where this active, outward extension is no kind of extramission. Olivi goes further than Merleau-Ponty, however, in coupling the active, outward extension of the illumination with being shaped by its \emph{terminus}, the circular or quadrangular vase, say. In illuminating a circular vase, the area illuminated is itself circular. The shape of the area illuminated is constituted by the shape of the object illuminated. The illumination is ``in conformity with the figure of the object which receives it and confines it.'' This is meant to be an analogy for how the perceptual act formally assimilates to its object. Extension and absorption are linked. They are a kind of procession and return. Olivi thinks that the perceptual act only assimilates to its object thanks to the activity of the perceptual soul.

Extension and absorption, a kind of procession and return, is important, so it is perhaps worth a brief digression on a detail of the passage that we have so far glossed over. The perceiver's gaze, in being fixed on its object, a circular vase, say, absorbs the object intentionally to itself. It is only in intentionally absorbing the object of perception that the perceptual act assimilates to its object. Moderns should resist the temptation to understand the qualifier ``intentionally'' in terms of the notion of intentionality derived from \citet{Brentano:1874aa} (on the historical development of the concept of intentionality see \citealt{Sorabji:2003fk}; on Olivi's role in the development of intentionality in late Scholasticism see \citealt[chapter 2]{Pasnau:1997aa}). A sensible form inhering in a body, the whiteness inhering in a circular vase, say, has natural existence in that body. Part of the point of the qualifier is to deny that the perceived sensible form has natural existence in the perceptual act. In part, then, the point of the qualifier is to rule out a position like the one William Crathorn will later endorse in his commentary on Lombard’s Sentences: ``A soul seeing and intellectively cognizing color is truly colored,'' \emph{Quaestiones super librum sententiarum} q. 1 concl. 7 \citep[288]{Pasnau:2002pb}. However, not only does Olivi deny natural existence to the sensible form in the perceptual act, he denies, as well, its real existence. This prompts \citet[67]{Pasnau:1997aa} to remark that with Olivi, there is ``movement toward making intentionality mysterious.''

Moreover, according to Olivi, the intentional existence of the object in the perceptual act---in virtue of which it assimilates to that object and so becomes like it, if not naturally like, in the manner of Crathorn---involves the perceptual power's virtual presence to that object (on the meaning of \emph{virtualis} in late Latin and Olivi specifically see \citealt[172--173]{Pasnau:1997aa}). Specifically, it is because the power is virtually present to its object that that object comes to exist intentionally in the actualization of that power: 
\begin{quote}
	A power can be present to something either essentially or virtually. This is to say that it can be present to something in such a way that its essence really is beside that thing, or in such a way that the gaze (\emph{aspectus}) of its power is so efficaciously directed to the thing that it, as it were, really touches the thing. If the power is not present to its object or recipient (\emph{patienti}) in this second way, it cannot act, even if it were present to it by its essence or according to the first way. The visual power is present to a thing that is seen from a distance in this [second] way. \dots\ This [kind of] presence suffices for an act of seeing. (Olivi, \emph{Quaestiones in secundum librum Sententiarum}, q. 58 486--487; \citealt[151--152]{Toivanen:2013ul})
\end{quote}
In speaking of a power's presence to its object as opposed to the object's presence to the power, Olivi is emphasizing the active, outer-directed nature of that power. If a power is essentially present to an object, then the power and the object are contiguous, ``its essence really is beside that thing,'' and there is a real connection between them akin to the perception by contact involved in touch. In contrast, if a power is virtually present to an object, then the object and the power are not contiguous but are at a distance from one another. Moreover, there is no real connection between the object and the power whose act contains it. Virtual presence is a necessary condition for visual perception. It is only by the visual power being virtually present to an object that seeing that object may formally assimilate to it. Virtual presence is also a sufficient condition. The virtual presence of the visual power to an object suffices for its extensive apprehension. 

Like intentional existence, the virtual presence of a power to an object contrasts with, not only the natural existence of that object in that power's act, but its real existence as well. Moreover, while the presence of the object in the natural environment may be required for its perception, it is not among the efficient causes of perception. But if what is intentionally absorbed by the perceptual act lacks both natural and real existence, and the object of perception in no way acts upon the perceiver, one may well wonder how, exactly, it may shape that act, even intentionally, such that the perceptual act formally assimilates to its object.

To bring this out, consider the way the neo-Platonic analogy fails to support Olivi's extreme position. Indeed, attending to its details, reveals a striking \emph{aporia}. The object of illumination, the illuminated circular vase, say, receives and confines that illumination. In receiving and confining the illumination the illuminated area takes on the shape that it does. In receiving and confining the illumination the circular vase resists that illumination. It obstructs that illumination and so casts a shadow. It is hard to understand how the spherical vase may confine, resist, and obstruct the activity of the illuminant without being a cause, or, at least, a countervailing force. Of course, it is the source of the illuminant that generates the illumination, but the illuminated area takes on the shape that it does because the illuminated object resists the activity of the illumination insofar as it can. Kilwardby's doctrine that the soul's use of a body is limited by the passivities of matter (\emph{De Spiritu Fantastico} 99–100) was meant to address this kind of difficulty. However, the invocation of the neo-Platonic analogy just is Olivi's response. Olivi is drawing our attention to the fact that it is the source that generates the illumination and not the object illuminated. But that does not suffice to make the analogy consistent with taking the object of perception to be a terminative cause with all that that entails. Visual consciousness may extend to its object, but it must somehow come into conflict with it, if the subsequent absorption is to be so much as possible. 

How is the Peripatetic analogy (\emph{De Anima} 2 12 424\( ^{a} \)18--23), the ancient figure of the wax and seal, meant to be understood by Olivi's lights? It occurs at the point where Olivi spells out the consequences of the neo-Platonic analogy for perception. One curious feature of Olivi's treatment is the way that way that extension and absorption are transposed at this point. Whereas earlier in the passage Olivi speaks of the act's extension to its object, he now speaks of the ``formative absorption of the act towards the object''. And whereas earlier in the passage Olivi speaks of the act's absorption of the object by which the act assimilates to it, he now speaks of ``a certain signet-like and inward extension of the object''. I am uncertain of the significance of this transposition. Perhaps it is meant to emphasize the unity of extension and absorption. Extension is at once a formative absorption to the object, just as absorption is at once an inward extension of the object. Notice, on this hypothesis, the unity of extension and absorption only holds for extensive apprehension, the kind of extensive activity characteristic of perception, as opposed to a non-perceptual visual experience, such as a hallucination. In cases of hallucination, there is nothing to absorb. And so while such experiences may involve extensive activity, there is no subsequent absorption.

The \emph{aporia} involved in Olivi's use of the neo-Platonic imagery affects his treatment of the ancient figure of the wax and seal. Even if, in line with the neo-Platonic analogy, the visual power generates the perceptual act in conformity with the figure of the object which receives it and confines it, how are we to understand this reception and confinement? ``Because it is generated thus the act becomes a similitude and signet-like expression of the object.'' Perception formally assimilates to its object because it is generated thus. It only conforms with its object by being received and confined. But reception and confinement is naturally understood as arising in the face of a countervailing force, the upshot of a conflict between the perceptual act and its object that resists it insofar as it can. It is hard to understand how the presence in the natural environment of an object which is the \emph{terminus} of the perceptual act could determine the content of that act, even if the act is directed upon it, like a beam of light, without somehow coming into conflict with it. Somehow the \emph{terminus} must determine the content of the perceptual act without itself being a determinant. But how could that be?

The present worry is anticipated by Duns Scotus. Scotus at least presses a parallel point about the intellect in his \emph{Ordinatio} and on the same general grounds. Though Scotus does not name names, Olivi is clearly a target as he reproduces a number of arguments from Olivi's \emph{Sentences} commentary \citep[148]{Pasnau:1997aa}. Scotus concedes to Olivi that the object of the intellect could not be the complete cause of the intellectual act. However, Scotus insists that the object must play some causal role if the act of intellect is to be a likeness of it (\emph{Ordinatio} 1 3 3 4 n. 486). Generalizing, Scotus' idea is that the demands of formally assimilating to the object require that the object play an explanatory role inconsistent with being a terminative cause. And it is the application of this general idea to the case of perception that constitutes the present worry (on Scotus on Olivi see \citealt[chapter 4.4]{Pasnau:1997aa}, on Scotus on cognitive powers, both sensory and intellectual, see \citealt[chapter 3]{Tachau:1988aa}, \citealt[257--266]{Spruit:1994qq}, and \citealt{Cross:2014aa}, for a related worry see \citealt[174--175]{Pasnau:1997aa}).

Even if Olivi is wrong to deny that an object plays a causal role in its perception, he may be right in claiming that extension and absorption are linked. Indeed, Olivi was criticized precisely by holding fast to the link between extension and absorption, a kind of procession and return, and drawing out what that entails, namely, that the active, outward extension's coming into conflict with the object is what explains, in part, that object's subsequent absorption. The grasping hand only conforms to rigid, solid body by grasping it. The grasping hand extends its grip until it can no more and so conforms to the body's contours. It is only thanks to the activity of the hand and the resistance that it encounters that the perceiver's haptic experience formally assimilates to the overall shape and volume of the object grasped. It is the force of the hand's activity coming into conflict with the self-maintaining forces of the object grasped that makes possible the presentation of that object in haptic experience and its formal assimilation to that object, understood as a mode of constitutive shaping. The grasped object plays an explanatory role, inconsistent with being a mere terminative cause, in the conflict with the hand's grasp that discloses it. If perception's formal assimilation to its object, understood as a mode of constitutive shaping, is the basis of its objectivity, that is only so because of the explanatory priority of its object, an explanatory priority inconsistent with being a terminative cause.

% section the_truth_in_extramission (end)

\section{Looking} % (fold)
\label{sec:looking}

I turn, and look, and see an ancient chestnut tree. It is one of the ancient chestnut trees replanted in Greenwich Park when Charles \textsc{ii} had the park redesigned in the 1660s. An organism of impressive size and age presents itself. The majority of its burrs remain on the tree and are brighter green than the surrounding foliage. It is early evening, and the light is long and golden. The light both articulates the fine texture of the bark and sets off the overall flow of the trunk in dramatic relief. Despite its manifest strength and solidity, the twisted trunk appears to be flowing in a wave-like form. I come to realize that I am witnessing an organic process, the growth of the trunk, occurring so slowly as to appear, from within my limited temporal perspective, to be frozen, static. The difference in the scale of our lives is striking. For a moment, it induces in me a kind a kind of temporal vertigo.  Just as a radical difference in spatial scale can be vertiginous---think of how small one can feel when viewing the Milky Way---a radical difference in temporal scale can be vertiginous as well. The scale of its life and the strength manifest in centuries of growth make the sweet chestnut tree a fit object of awe. I find myself musing that in a different cultural context, perhaps one more prone to animism, the tree might reasonably be reckoned a god. 

In looking at the ancient chestnut tree, I do so from across the park. My gaze perceptually penetrates that space until it encounters the ancient tree. The tree's surface is the site of visual resistance. Perceptually impenetrable, it determines a visual boundary through which, and in which, nothing further may appear. The tree is opaque to a significant degree. Its opacity consists in its resistance to my gaze. The illuminated air between, by contrast, being transparent, is perceptually penetrable. One can see through it, and in it. Thus a scrub brush can appear in the water of a bath, and a cherry tree can appear through a window. Appearing through a medium does not require that the object be embedded in that medium the way appearing in does, though it is consistent with the object being so embedded at least if the perceiver is as well. Thus, it is through the illuminated air that the ancient chestnut tree is disclosed to me in sight. Looking involves the perceiver's gaze coming into conflict with what is perceptually impenetrable. 

\citet{Broad:1952kx} describes vision as prehensive and saltatory. It is prehensive insofar as vision involves the presentation of its object in the awareness afforded by visual experience. It is saltatory insofar as this awareness seems to leap the spatial gap between the perceiver and the object. There are two separable elements to Broad's conception of saltatoriness. The first is simply the frank admission that vision is a kind of perception at a distance, that the objects of visual awareness are located at a distance from the perceiver. That much is unexceptional. The second is a phenomenological claim, that visual awareness seems to leap the spatial gap between the perceiver and the object of perception. For visual awareness to leap the spatial gap would be for the objects of visual awareness to be confined to a remote location.

Two years after the appearance of ``Some elementary reflections on sense-per\-cep\-tion'', \citet[518]{Jonas:1954aa} will deny that vision is saltatory in Broad’s sense, and it is the second element of Broad's conception that he takes exception to and not the first: ``in sight the object faces me across the intervening distance, which in all its potential `steps' is included in the perception''. Broad is right to emphasize the distal character of the objects of vision, but his description of vision as saltatory is inapt since it fails to heed the perceptual penetrability of the intervening medium. Vision would leap the gap between the perceiver and the distal color if the object of visual awareness were confined to the remote spatial region where that color is instantiated. However, vision is not so confined and so does not leap the gap between the perceiver and distal color. Rather, by means of it, the perceiver may peer through the intervening medium, in all its potential steps, and encounter objects facing them across the intervening distance, if the medium is transparent at least to some degree. In the course of an otherwise astute and insightful comparative phenomenology of the senses, Broad is misled, at this point, by overlooking the active, outer-directed phenomenology of vision. Broad, in effect, overlooks the truth in extramission.

This psychological stance may be sustained, in the Peripatetic fashion, by a capacity to act. Looking, while not a passive power, may be less than fully active. Looking may be a psychological stance sustained, at a minimum, by the potential to act in visually relevant ways, to alter one's visual perspective on the natural environment to better bring into view distal aspects of that environment, but only on a particular understanding of that potentiality. Looking at the ancient chestnut tree may involve, at a minimum, a preparedness to act in such visually relevant ways, but such preparedness requires vigilance. In looking at the ancient chestnut tree, I maintain vigilance over the tree and its visually manifest aspects. Being thus vigilant, being prepared to act in visually relevant ways, remains a stance that I must actively sustain. So the characteristic activity that sustains the psychological stance may be a potentiality, but the relevant sense of potentiality involves a preparedness to act in a way that itself requires activity to sustain, a kind of perceptual vigilance.

Looking may be a psychological stance, sustained by a characteristic activity, where the perceiver opens themselves up, in a directed manner, to visually experience distal aspects of the natural environment, but that stance is itself an activity. In maintaining perceptual vigilance, I open myself up to the visible. My gaze, that the tree resists insofar as it can, is something I direct at the tree. Looking through a window, or into a fish tank, or across a park is something that the perceiver does. Looking at the tree, gazing upon it, remains something that I do, even if in seeing the tree I undergo an experience caused in me, at least in part, by the tree itself.

Looking, so conceived, may not be a simple, spiritual act of the immaterial soul as Olivi maintains, but its outward, extensive activity remains something that the perceiver does independently of any visible object it may encounter. In opening their eyes, the perceiver opens themselves up to visually experiencing the natural environment, and that is something they do independently of whatever they encounter in so doing. However, accommodating this insight, if it is one, does not require the object of perception to be a terminative cause. In openning themselves up, in a directed manner, to visually experiencing distal aspects of the natural environment, the content of their perception is determined by what they encounter in so looking in a manner inconsistent with the object of perception being a mere terminative cause.

Looking, understood as a psychological stance sustained by characteristic activity, is an outward gaze, a looking into the distance, an outer-directed opening up to the visible. It can sometimes happen, if circumstances are propitious, that in looking outward, aspects of the natural environment, facing us from across the intervening distance, are presented to us in our visual experience. 

% section looking (end)

\section{Sympathy and Visual Presentation} % (fold)
\label{sec:sympathy_and_visual_presentation}

I turn, and look, and see an ancient chestnut tree. In so doing, I direct my gaze across the park. I look through the illuminated space, a space perceptually penetrated by my gaze, until I can no more. It is the resistance to my looking, my visual encounter with the perceptually impenetrable, that presents opaque objects arrayed in the distal environment. The ancient chestnut tree resists my visual activity. The ancient chestnut tree prevents me from seeing further. I can see nothing in it or through it. However, not all limits to my gaze are external. There are internal limits to how far I may look into the distance. Other perceivers posses the capacity to look further than I can. So how is it possible for an experienced limit to my visual activity to disclose the perceptually impenetrable tree? If the visual presentation of the perceptually impenetrable is due to the operation of sympathy, then we have the basis of an answer. It is only when I experience the tree's limit to my visual activity, its resistance to my gaze, its perceptual impenetrability, as a sympathetic response to a countervailing force, my gaze encountering an alien force that resists it, one force in conflict with another, like it yet distinct from it, that the perceptually impenetrable body discloses itself to visual awareness.

To get a sense of this, compare David Katz's description of the way that the appearance of surface color contrasts with the appearance of spectral color:
\begin{quote}
	The paper has a surface in which the colour lies. The plane on which the spectral color is extended in space before the observer does not in the same sense possess a surface. One feels that one can penetrate more of less deeply \emph{into} the spectral color, whereas when one looks at the colour of a paper the surface presents a barrier beyond which the eye cannot pass. It is as though the colour of the paper offered resistance to the eye. We have here a phenomenon of visual resistance which in its way contributes to the structure of the perceptual world as something existing in actuality. \citep[8]{Katz:1935qv}
\end{quote}
The phenomenon of visual resistance contributes to the structure of the perceptual world as something existing in actuality. And it does so, or so I claim, by being a necessary precondition for the sympathetic presentation of what resists the perceiver's gaze. Katz's discussion also nicely brings out how, from among the many determinate forms of visual resistance, there is a distinctly chromatic form of visual resistance at work in the contrasting appearances of surface and spectral color.

Sympathy, here, is not an emotional response to others, a kind of fellow-feeling, akin to compassion or pity. Rather, the notion of sympathy being invoked is closer, instead, to the notion at work in Stoic physics and Plotinus, if more abstract and not at all reliant on on their vitalistic metaphysics (on sympathy in Stoic physics see \citealt{Brouwer:2015ee}, in Plotinus see \citealt{Emilsson:2015wf}). The present approach thus contrasts with Whitehead’s \citeyearpar{Whitehead:1978zr} who both explains perceptual prehension partly in terms of sympathy and embraces the association with emotion thus leading him to paradoxically portray perception as an outgrowth of blind emotion.

Despite philosophers' penchant for limiting their visual examples to opaque bodies, such as Moore's \citeyearpar{Moore:1903uo} blue bead or Price's \citeyearpar{Price:1932fk} red tomato, not all \emph{visibilia} are opaque and not all are bodies. Can the account of the sympathetic presentation in vision of opaque bodies be extended to, at least, non-opaque things? Is the principle of sympathy operative in the presentation of the visible more generally?

In \emph{De Sensu}, Aristotle observes that transparency comes in degrees. The transparent offers insufficient visual resistance to determine a perceptually impenetrable boundary. But offering insufficient visual resistance to determine a perceptually impenetrable boundary is consistent with offering visual resistance nonetheless. Something is perfectly transparent if it offers no visual resistance to sight. Something is imperfectly transparent if it offers visual resistance to sight but not sufficient to determine a perceptually impenetrable boundary. From perfect transparency, as we approach the limit of perceptual penetrability, opacity, there is a range of states of imperfect transparency ordered by declining degrees of perceptual penetrability. 

The illuminant is a perceptual medium. We may see an opaque body through or in the illumination. Whereas physical media answer to the demands of being a causal intermediary, perceptual media answer to the demands of perceptual accessibility. Light does not require physical media in which to propagate in the way that sound waves do. As the Michelson–Morley experiment of 1887 went some way toward showing, there is no Luminiferous aether. But the illuminated air may be a perceptual medium, nonetheless. Moreover, not only are perceptual media themselves perceptible, but they are perceptible in a certain way. Specifically, they are not perceptible in themselves, but owe their perceptibility to other things which are perceptible in themselves, the objects the perceptual media make perceptually accessible. So the illuminant is visible, though not visible in itself, but owes its visibility to the objects that it illuminates. One sees the brightness of a pantry, not in itself, but by seeing the brightly lit objects arranged in it. This is the way in which the perceptually penetrable presents itself to the perceiver's gaze.

The more the perceptually penetrable resists the perceiver's gaze, the more visible in its own right it becomes and so loses, to that degree, the capacity for other things to be perceived in it, or through it. Visual resistance can take many forms. For example, the determinate kind of visual resistance offered by a perceptually penetrable thing, such as a liquid mass, may consist in its possessing a volume color. A volume color pervades the perceptually penetrable mass, and that liquid mass has that color, independently of the colors of the things arrayed in it, or seen through it \citep[though see][]{Mizrahi:2010aa}. If the liquid mass is sufficiently perceptually penetrable, seeing the colors of things arrayed in it may be within the bounds of normal human color constancy. That is, one may see a red bead in a yellow liquid and that bead may be seen to be red, though, of course, looking the way a red thing would when seen through a yellow liquid. The red bead will look to be red, and the same shade of red, when seen through a clear liquid, though, of course, it will look another way. In moving from the yellow liquid to the clear, the red bead's appearance changes but the bead does not appear to change color. There are limits, however, to the normal human color constancy. If the liquid is strongly enough colored, if it offers sufficient visual resistance in that way, this will erode the perceiver's ability to visually recognize the determinate shade of the bead, or even that it is red. Volume color is not the only form of visual resistance offered by otherwise perceptually penetrable media. As Katz observed, spectral color also offers visual resistance. And refractions, reflections, specular highlights, shadows, all contribute, in determinate ways, to the visual resistance of the imperfectly transparent.

The perfectly transparent, insofar as it can be seen at all, is visually presented by the objects seen in it or through it. Its visibility is entirely parasitic on the visibility of the objects it enables. Insofar as the perceptually impenetrable is presented in sight as a sympathetic response to the experienced limit to the perceiver's gaze, and the perfectly transparent medium is thereby presented, the principle of sympathy makes possible the presentation, in vision, of the perfectly transparent. The imperfectly transparent, by contrast, offers visual resistance at least to some degree, but not to a degree sufficient to determine a perceptually impenetrable boundary. To the degree that it manifestly resists perceptual penetration, it is possible to sympathetically present it in visual experience. Think of the way in which the volume color or refraction of an imperfectly transparent medium may present that medium in our visual experience of it. However, the more visible in its own right the imperfectly transparent becomes, the more it erodes the sympathetic presentation of objects arrayed in that medium. The more we see visible features of the illuminated media had independently of the objects that it illuminates, the less capable we are of seeing through it or in it. Illumination may reveal the latent visibility of things, but if it is sufficiently strong, it may blind us to the scene. Perceptual media, in calling attention to themselves, erode the sympathetic presentation of distal objects they otherwise make possible.

So we have the following argument by cases. The visible exhaustively divides into the perceptually impenetrable and the perceptually penetrable. The perceptually penetrable is either perfectly perceptually penetrable, offering no visual resistance, or imperfectly penetrable, offering visual resistance to some degree. The operation of sympathy suffices to explain the visual presentation of the perceptually impenetrable. Moreover this explanation suffices, as well, for the visual presentation of the perfectly penetrable, as we have explained. Sympathy explained as well not only the presentation of the imperfectly penetrable insofar as other objects may be sympathetically presented in it, or through it, but also the respects in which it is visible in its own right and the way that this erodes the sympathetic presentation of objects seen in it, or through it. So the operation of sympathy suffices for the presentation of the visible, in sight, quite generally.

We began by explaining the visual presentation of an opaque body in terms of sympathy. Since the objects of sight are not limited to opaque bodies, this raised the question whether sympathy operates in visual presentation quite generally. The following worry might arise about the argument so far: While we have explicitly addressed the visual presentation of non-opaque things, we have failed to explicitly address the visual presentation of non-corporeal things, such as events and processes. However, perceptual impenetrability does not merely pertain to the surfaces of opaque bodies. A flame, should the fire be burning intensely enough, may be perceptually impenetrable, obstructing the view of other \emph{visibilia}. The general point is that the way in which we have characterized the visible, in terms of degrees of perceptual penetrability, is equally applicable to visible objects of distinct ontological categories. Perceptual penetrability applies equally to corporeal and non-corporeal things and so does not preclude the visual presentation of events and processes.

In sympathetically disclosing the ancient chestnut tree, my visual experience absorbs that tree and is constitutively shaped by it. The conscious character of seeing the tree is constituted, in part, by its bright green burs and the wave-like form of its trunk sympathetically presented to my partial perspective on that tree in the given circumstances of perception. What it is like for me to see the tree depends upon and derives from, at least in part, what the tree is like, at least in visible respects. Visual experience formally assimilates to its object, relative to the perceiver's partial perspective, as a consequence of being constitutively shaped by that object as presented to that perspective, a constitutive shaping made possible by the sympathetic presentation of that object in visual experience. Constitutive shaping of visual experience by its object is a ``communion'' with that object (see Ardley's \citeyear{Ardley:1958aa} unjustly neglected essay). In undergoing that experience the perceiver is united, in a way, with the object of their perception. This unity explains in part, the similarity between the visual experience and its object. The formal assimilation of visual perception to its object, at least relative to the perceiver's partial perspective, is the effect of constitutive shaping, and thus its conscious character depends upon and derives from, at least in part, the visible character of the object seen.

The unity of visual presentation is explanatorily prior to the operation of sympathy. The visual presentation of distal aspects of the natural environment is not being constructed from elements and principles understood independently of their visual presentation, rather the unity of the perceiver and the distal aspects of the natural environment is presupposed, and sympathy merely analytically explicates the intelligible structure of this presupposed unity. The analytic approach to sensory presentation is comparable to Frege's approach towards thought, at least at certain stages of his career, on certain interpretations (see, for example, \citealt[essays 7 and 9]{Travis:2011qd}). Frege begins with a unity, a truth-evaluable thought, and discerns what intelligible structure it must display. Beginning with the thought, Frege analyzes or decomposes that thought into constituent elements that must be present and principles that must be operative if that thought is to be so much as truth-evaluable (which is not say that there is a unique such decomposition). Frege's position thus contrasts with recent discussions of the problem of the unity of the proposition (compare \citealt{King:2007ad}, \citealt{Gaskin:2008aa}, \citealt{Soames:2010qq}, and \citealt{King:2014ls}). The problem of the unity of the proposition simply does not arise for Frege, since he does not begin with independently understood elements and principles and tries to construct thoughts on their basis. Rather the unity of thought is explanatorily prior to the intelligible structure it must display if it is to be so much as truth-evaluable. Similarly, on the analytic approach, the unity of sensory presentation is explanatorily prior to the intelligible structure it must display if it is so much as possible.

Not only does sympathy only operate within a unity, but that unity is reducible to no other thing. Visual presentation is an irreducible unity. If sensory presentation is a distinctive kind of unity, a ``communion'' with its object, then visual presentation is more distinctive still. Insofar as visual presentation is governed by the principle of sympathy, it is a mode of being with. Turning, and looking, and seeing the ancient chestnut tree is a way of being with that tree. Sartre's overly aggressive conception of the look, in \emph{L'Être et le néant: Essai d'ontologie phénoménologique}, perhaps blinds him to this possibility. Sartre fails to see how the look's coming into conflict with its object may be the means of the latter's sympathetic presentation to the former. (See \citealt[chapter 5]{Jay:1994aa}, especially 287, where he remarks that Heidegger's conception of \emph{mitsein} was, perhaps, too irenic for Sartre). Visual presentation is a way of being with that which resists the perceiver's gaze.
% section sympathy_and_visual_presentation (end)

\bibliographystyle{plainnat}
\bibliography{Philosophy}

\end{document}